Approximately Exact Calculations for Linear Mixed Models

### 2015-05-19 ms.tex
edits to Sec 2.1, 2.2

As \cite{hodges:2013} further observes, if $\beta$ is given an improper flat prior and $\sigesq$ and $\sigssq$ are given conjugate priors --- say $\sigesq \sim \IG(\alpha_e,\beta_e)$ and $\sigssq \sim \IG(\alpha_s,\beta_s)$ --- then \begin{equation*}
  -(\alpha_e+1) \log\sigesq - \beta_e/\sigesq -(\alpha_s+1) \log\sigssq - \beta_s/\sigssq
\end{equation*}
is added to \eqref{eq:reexpress} to yield
the log posterior
\begin{equation}
\label{eq:logpost}
  \begin{split}
  \log\pi(\sigesq,\sigssq\g y) &=
  B - \frac{n_e + 2\alpha_e + 2}{2}\log(\sigesq) -
    \frac{y^t \Gamma_c \Gamma^t_c y + 2\beta_e}{2\sigesq}\\
    &\qquad\qquad - \frac{2\alpha_s + 2} {2}\log(\sigssq) - \frac{2\beta_s}{2\sigssq}
    - \frac{1}{2} \sum_{j=1}^{s_z} \left[ \log(\ass) + \frac{\hat v_j^2}{\ass}\right].
  \end{split}
\end{equation}
Equations~\eqref{eq:reexpress} and~\eqref{eq:logpost} can both be written as a sum of multiples of logs and inverses of linear combinations $\abss$, as in \eqref{eq:sum}, where the summands with $j=s_z+1$ and $j=s_z+2$ are for the terms involving only $\sigesq$ and $\sigssq$, respectively.
\begin{equation}
\label{eq:sum}
  -\frac{1}{2} \sum_{j=1}^{s_z+2}\left[ c_j \log(\abss) + \frac{d_j}{\abss}\right]
\end{equation}
where
\begin{itemize}[label=---]
\item for $j = 1, \dots, s_z$,
  \begin{align*}
    a_j &> 0\ \text{and is derived in the Appendix}\\
    b_j &=1\\
    c_j &=1\\
    d_j &= \hat v_j^2\ \text{and is derived in the Appendix}
  \end{align*}
\item for $j = s_z +1$,
  \begin{align*}
    a_j &= 0\\
    b_j &= 1\\
    c_j &= \begin{cases}
                 n_e & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
                 n_e + 2\alpha_e + 2 & \text{for the $\log$ posterior in \eqref{eq:logpost}}
              \end{cases}\\
    d_j &= \begin{cases}
                 \frac{y^t \Gamma_c \Gamma^t_c y}{\sigesq} & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
                 \frac{y^t \Gamma_c \Gamma^t_c y + 2\beta_e}{\sigesq} & \text{for the $\log$ posterior in \eqref{eq:logpost}}
              \end{cases}
  \end{align*}
\item for $j = s_z+2$,
  \begin{align*}
    a_j &= 1\\
    b_j &= 0\\
    c_j &= \begin{cases}
      0 & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
      2 \alpha_s + 2 & \text{for the $\log$ posterior in \eqref{eq:logpost}}
    \end{cases}\\
    d_j &= \begin{cases}
      0 & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
      2 \beta_s & \text{for the $\log$ posterior in \eqref{eq:logpost}}.
    \end{cases}
  \end{align*}
\end{itemize}
Our derivation proceeds from \eqref{eq:sum}.  We use $f$ to denote the target function generically, either~\eqref{eq:reexpress} or~\eqref{eq:logpost}.  When the target function is~\eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$ and the upper limit of the sum in \eqref{eq:sum} is $s_z+1$.

@@

\section{Satisfying the Desiderata}
\subsection{Partial Derivatives Determine Lines}
The partial derivatives of $\logRLss$ can be calculated from (\ref{eq:logpost}):
\begin{subequations}
\label{eq:partials}
\begin{equation}
\label{eq:drlds}
  \frac{\partial\logRLss}{\partial\sigssq}
  = -\frac{1}{2} \sum_j
        \left[ \frac{a_jc_j}{\abss} - \frac{a_j d_j}{(\abss)^2} \right]
        = -\frac{1}{2} \sum_j
           \frac{a_j \left( a_jc_j\sigssq + b_jc_j\sigesq  - d_j\right)}{(\abss)^2}\\
\end{equation}
and
\begin{equation}\label{eq:drlde}
  \frac{\partial\logRLss}{\partial\sigesq} =
    -\frac{1}{2} \sum_j \left[ \frac{b_jc_j}{\abss} - \frac{b_j d_j}{(\abss)^2} \right]
    = - \frac{1}{2} \sum_j \frac{b_j \left( a_jc_j\sigssq + b_jc_j\sigesq  - d_j\right)}{(\abss)^2}.
\end{equation}
\end{subequations}

We work with one term in (\ref{eq:partials})'s summations at a time; that is, one $j$ at a time.  For $j=1, \dots, s_z$, the $j$'th terms in (\ref{eq:partials}) differ  by a multiplicative constant $a_j/b_j = a_j$; they have the same sign as each other and the same sign as $(a_jc_j\sigssq + b_jc_j\sigesq  - d_j) = (a_j\sigssq + \sigesq  - d_j)$, which determines a line $\sigssq = d_j/a_j - \sigesq/a_j$ --- call it the $j$'th line --- in the first quadrant of the $(\sigssq,\sigesq)$ plane.  The $j$'th line has positive intercept $ d_j/a_j$ and negative slope $-1/a_j$.  For $j=s_z+1$, $a_j=0$, so $\eqref{eq:drlds} = 0$ and \eqref{eq:partials} determines a vertical line at $\sigesq = \sigma_e^{2*} \equiv d_{s_z+1} / c_{s_z+1}$.  For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$.

@@

Figure~(\ref{fig:bigboundingbox}) shows five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.  The largest intercept of those lines on the $\sigssq$ axis and the largest intercept on the $\sigesq$ axis determine a rectangle $B_1$ whose outline is marked in bold.  At any point above $B_1$, e.g., $p_1$, the partial derivatives \eqref{eq:partials} are either negative or, for $j=s_z+1$, zero.  Therefore, $\logRL(\widetilde{p_1}) \ge \logRL(p_1)$.  Similarly, $\logRL(\widetilde{p_2}) \ge \logRL(p_2)$ and $\logRL(\widetilde{p_3}) \ge \logRL(p_3)$.  That is, for points above and/or to the right of $B_1$, $\logRLss$ can be increased by moving down and/or to the left as far as $B_1$'s boundary.  Therefore \mrle\ must lie on or inside $B_1$ and there can be no maxima of $\logRLss$ in $B_1^c$.  $B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.  However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$.  This paper presents a computational method guaranteed not to miss \mrle\ and which evaluates $\logRLss$ everywhere to within a specified tolerance inside a box $B$ satisfying desideratum \textbf{D1}.  First, though, we pause to note that the region outlined in bold in Figure~(\ref{fig:smallboundingbox}) is a subset of $B_1$ that must also contain \mrle\ for the reasons given above.  But it is not rectangular, hence less convenient than $B_1$, so we don't pursue it further.

@@

$B$ is determined by its intercepts $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ on the $\sigesq$ and $\sigssq$ axes respectively.  Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis and let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$. Points such as $q_1$ which satisfy $\sigesq \le \sigma_e^{2*}$ and $\sigssq \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the left of the $s_z+1$ line --- are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are negative for $j \in \{1, \dots, s_z, s_z+2\}$ but nonnegative for $j=s_z+1$.  Let $\logRL_j(\cdot)$ denote the $j$'th term of \eqref{eq:sum} evaluated at $(\cdot)$.  
\begin{equation}
	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2})
\end{equation}
which is larger than $M$ if
\begin{equation}
\label{eq:q1}
  \begin{split}
    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
  \end{split}
\end{equation}
Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.  Thus $\widetilde{\sigma_s^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q1} is less than the r.h.s.~of \eqref{eq:q1}, and  \textbf{D1} is satisfied.

@@

Let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$.  For any point $q_4$ having $\sigssq \le \sigma_s^{2*}$ and $\sigesq \ge \widetilde{q_4}$, the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_z+1\}$  but nonnegative for $j=s_z+2$.  
\begin{equation*}
	\logRL(\sshat) - \logRL(q_4) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
\end{equation*}
which is larger than $M$ if
\begin{equation}
\label{eq:q4}
  \begin{split}
    \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
      &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
      &= L - M + \frac{1}{2} \left[ c_{s_z+2}\log\sigma_s^{2*} + \frac{d_{s_z+2}}{\sigma_s^{2*}}\right].
  \end{split}
\end{equation}
Examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, and for $j \ne s_z+2$,\\ $\lim_{\sigesq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.  Thus $\widetilde{\sigma_e^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q4} is less than the r.h.s.~of \eqref{eq:q4}, and  \textbf{D1} is satisfied.

Combining the requirements for points like $q_1$, $q_2$, $q_3$, and $q_4$, we satisfy desideratum \textbf{D1} for all points in $B^c$ by choosing $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ sufficiently large. 
%enough to satisfy $\logRL(q_1^*) < L-M$ and choosing $\sigma_s^*$ large enough to satisfy $\logRL(q_2^*) < L-M$ and $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.  In addition, because $\logRL$ has no maxima outside of $B_1$, it also has no maxima outside of $B$.  As already mentioned, if there is an $h$ line, then $q_1$ would have to be treated like $q_2$ and $q_3$.

@@

\subsection{Lines Determine Bounds within Boxes}
\begin{figure}
	\centering
	\includegraphics[width=.5\linewidth]{figs/boxb.pdf}
	\caption{For a box $b$,  the minimum and
	maximum within $b$ of $\logRL_j$
	occur at either the lower left corner, the upper right  corner, or
	on the line.
}	\label{fig:boxb}
\end{figure}
Next we consider the relationship between the $\logRL_j$'s and boxes as depicted in Figure~\ref{fig:boxb}, which shows a box $b$ and three lines.  (Though we haven't drawn $v$ or $h$ lines, the argument for them is similar to the argument for the lines we have drawn.)  For lines such as $j=2$ that lie below $b$ the partial derivatives of $\logRL_j$ are negative, so the maximum and minimum of $\logRL_j$ within $b$ are attained at the lower left and upper right corners respectively.  The situation is reversed for lines like $j=3$ that lie above $b$: the partial derivatives are positive so the maximum and minimum are attained at the upper right and lower left corners respectively.  For lines like $j=1$ that pass through $b$ the minimum is attained at either the upper right or lower left corner while the maximum is attained on the line.

@@

\section{An Algorithm}
With \textbf{D1} and \textbf{D2} the following  algorithm will evaluate $\logRL$ to arbitrary accuracy everywhere within a box $B$.
\begin{enumerate}
\item Specify constants \maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$.
	\begin{enumerate}[label=(\alph*)]
	\item \maxit\ is the maximum number (could be $\infty$) of iterations of
		the algorithm's loop.
	\item We will not carefully analyze regions of the plane where
		$\logRLss < \logRL(\sshat) - M$.  ($M$ could be $\infty$.)
	\item We will evaluate $\logRL$ to within an accuracy of $\epsilon$ (could be 0) inside $B$
	         unless evaluation is stopped by one of the other criteria.
	\item We will not distinguish values of $\sigesq$ separated by less
		than $\delta_e$ (could be 0) nor values of $\sigssq$ separated by
		less than $\delta_s$ (could be 0).  Separation may be specified
		in either absolute or relative terms. (I.e.\ we look at either the difference
		in $\sigesq$ or $\log\sigesq$ (or $\sigssq$ or $\log\sigssq$) from one
		side of the box to the other.)
	\end{enumerate}
	\maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$ are constants that
	control the running of the algorithm.  Their meaning may become more
	clear as we explain the algorithm.
\item With $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ --- see \eqref{eq:lmm}
	and \eqref{eq:rll} for definitions --- compute $\{a_j, b_j, c_j, d_j\}$ for $j=1,
	\dots, s_z, v, h$.
\item Create two lists of boxes: active and inactive.  Both lists are initially empty.
\item Specify a box $B$ as the sole entry of the active list.  $B$ could be,
	but doesn't have to be, a box satisfying desideratum \textbf{D1}.
\item Either set $L = -\infty$ or evaluate $\logRL(p_i)$ at a few points
	$p_1, \dots p_k$ and set $L = \max_i\logRL(p_i)$.  $L$ is our current
	lower bound on $\logRL(\sshat)$.
\item While the active list is not empty or we have not reached \maxit\
	iterations
	\begin{enumerate}
	\item For each active box $b$:
		\begin{enumerate}
		\item if $U^b < L-M$ move $b$ to the inactive list.
		\item if $U^b - L^b < \epsilon$ move $b$ to the inactive list
		\item if the horizontal extent of $b$ is less than $\delta_e$,
			move $b$ to the inactive list
		\item if the vertical extent of $b$ is less than $\delta_s$,
			move $b$ to the inactive list
		\item otherwise, $b$ is still active.  Divide $b$ into four subboxes.
			Remove $b$ and add the subboxes to the active list.
		\end{enumerate}
	\item Set $L = \max ( L, \max_{b\in \text{active list}}L^b )$.  $L$ is our new
		lower bound on $\logRL(\sshat)$.
	\end{enumerate}
\item Return the active and inactive lists.
\end{enumerate}

@@

\subsubsection{A log RL Analysis}
For a $\logRLssorig$ analysis there are 43 lines, as shown in Figure~\ref{fig:hmolines}.
Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ with the settings
\begin{equation*}
	\maxit=10; \hspace{1cm} \epsilon=5; \hspace{1cm}
	\delta_e=\log(10); \hspace{1cm} \delta_s=\log(10); \hspace{1cm} M=5
\end{equation*}
results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 10: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$.  We see that boxes are steadily transferred from the active to the inactive list and that $L$ increases monotonically.  After 10 iterations there is a total of 640 boxes, which are displayed in Figure~\ref{fig:hmo1}.

@@

\begin{figure}
	\centering
	\includegraphics[width=.45\linewidth]{figs/hmo_HH11_lines.pdf}
	\caption{The 43 lines, $j=1, 2, \dots, 42, v$, for the $\logRLorig$ analysis of the HMO data.
	              The $v$ line is dashed.}
	\label{fig:hmolines}
\end{figure}

>>>>

As \cite{hodges:2013} further observes, if $\beta$ is given an improper flat prior and $\sigesq$ and $\sigssq$ are given conjugate priors --- say $\sigesq \sim \IG(\alpha_e,\beta_e)$ and $\sigssq \sim \IG(\alpha_s,\beta_s)$ --- then \begin{equation*}
  -(\alpha_e+1) \log\sigesq - \beta_e/\sigesq -(\alpha_s+1) \log\sigssq - \beta_s/\sigssq
\end{equation*}
is added to \eqref{eq:reexpress} to yield
the log posterior
\begin{equation}
\label{eq:logpost}
  \begin{split}
  \log\pi(\sigesq,\sigssq\g y) &=
  B - \frac{n_e + 2\alpha_e + 2}{2}\log(\sigesq) -
    \frac{y^t \Gamma_c \Gamma^t_c y + 2\beta_e}{2\sigesq}\\
    &\qquad\qquad - \frac{2\alpha_s + 2} {2}\log(\sigssq) - \frac{2\beta_s}{2\sigssq}
    - \frac{1}{2} \sum_{j=1}^{s_z} \left[ \log(\ass) + \frac{\hat v_j^2}{\ass}\right].
  \end{split}
\end{equation}
Equations~\eqref{eq:reexpress} and~\eqref{eq:logpost} can both be written as a sum of multiples of logs and inverses of linear combinations $\abss$, as in \eqref{eq:sum}, where the summands with $j=s_z+1$ and $j=s_z+2$ are for the terms involving only $\sigesq$ and $\sigssq$, respectively and where we have dropped the irrelevant constant $B$.
\begin{equation}
\label{eq:sum}
  \logRLss = -\frac{1}{2} \sum_{j=1}^{s_z+2}\left[ c_j \log(\abss) + \frac{d_j}{\abss}\right]
\end{equation}
where
\begin{itemize}[label=---]
\item for $j = 1, \dots, s_z$,
  \begin{align*}
    a_j &> 0\ \text{and is derived in the Appendix}\\
    b_j &=1\\
    c_j &=1\\
    d_j &= \hat v_j^2\ \text{and is derived in the Appendix}
  \end{align*}
\item for $j = s_z +1$,
  \begin{align*}
    a_j &= 0\\
    b_j &= 1\\
    c_j &= \begin{cases}
                 n_e & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
                 n_e + 2\alpha_e + 2 & \text{for the $\log$ posterior in \eqref{eq:logpost}}
              \end{cases}\\
    d_j &= \begin{cases}
                 y^t \Gamma_c \Gamma^t_c y & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
                 y^t \Gamma_c \Gamma^t_c y + 2\beta_e & \text{for the $\log$ posterior in \eqref{eq:logpost}}
              \end{cases}
  \end{align*}
\item for $j = s_z+2$,
  \begin{align*}
    a_j &= 1\\
    b_j &= 0\\
    c_j &= \begin{cases}
      0 & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
      2 \alpha_s + 2 & \text{for the $\log$ posterior in \eqref{eq:logpost}}
    \end{cases}\\
    d_j &= \begin{cases}
      0 & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
      2 \beta_s & \text{for the $\log$ posterior in \eqref{eq:logpost}}.
    \end{cases}
  \end{align*}
\end{itemize}
Our derivation proceeds from \eqref{eq:sum}.  We use $f$ to denote the target function generically, either $\RLorig(\sigesq,\sigssq)$ or $\pi(\sigesq,\sigssq\g y)$.  When the target function is $\RLorig(\sigesq,\sigssq)$, $c_{s_z+2} = d_{s_z+2} = 0$ and the upper limit of the sum in \eqref{eq:sum} is effectively $s_z+1$.

@@

\section{Satisfying the Desiderata}
\subsection{Partial Derivatives Determine Lines}
The partial derivatives of $\logRLss$ can be calculated from (\ref{eq:sum}):
\begin{subequations}
\label{eq:partials}
\begin{equation}
\label{eq:drlds}
  \frac{\partial\logRLss}{\partial\sigssq}
  = -\frac{1}{2} \sum_j
        \left[ \frac{a_jc_j}{\abss} - \frac{a_j d_j}{(\abss)^2} \right]
        = -\frac{1}{2} \sum_j
           \frac{a_j \left( a_jc_j\sigssq + b_jc_j\sigesq  - d_j\right)}{(\abss)^2}\\
\end{equation}
and
\begin{equation}\label{eq:drlde}
  \frac{\partial\logRLss}{\partial\sigesq} =
    -\frac{1}{2} \sum_j \left[ \frac{b_jc_j}{\abss} - \frac{b_j d_j}{(\abss)^2} \right]
    = - \frac{1}{2} \sum_j \frac{b_j \left( a_jc_j\sigssq + b_jc_j\sigesq  - d_j\right)}{(\abss)^2}.
\end{equation}
\end{subequations}

We work with one term in (\ref{eq:partials})'s summations at a time; that is, one $j$ at a time.  For $j=1, \dots, s_z$, the $j$'th terms in (\ref{eq:partials}) differ  by a multiplicative constant $a_j/b_j = a_j$; they have the same sign as each other and the same sign as $(a_jc_j\sigssq + b_jc_j\sigesq  - d_j) = (a_j\sigssq + \sigesq  - d_j)$, which determines a line $\sigssq = d_j/a_j - \sigesq/a_j$ --- call it the $j$'th line --- in the first quadrant of the $(\sigssq,\sigesq)$ plane.  The $j$'th line has positive intercept $ d_j/a_j$ and negative slope $-1/a_j$.  For $j=s_z+1$, $a_j=0$, so $\eqref{eq:drlds} = 0$ and \eqref{eq:partials} determines a vertical line at $\sigesq = \sigma_e^{2*} \equiv d_{s_z+1} / c_{s_z+1}$.  For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$, while if the target function is \eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$ and the term for $j=s_z+2$ effectively vanishes.

@@

Figure~(\ref{fig:bigboundingbox}) shows five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.  The largest intercept of those lines on the $\sigssq$ axis and the largest intercept on the $\sigesq$ axis determine a rectangle $B_1$ whose outline is marked in bold.  At any point above $B_1$, e.g., $p_1$, the partial derivatives \eqref{eq:partials} are either negative or, for $j=s_z+1$, the partial derivative \eqref{eq:drlds} is zero.  Therefore, $\logRL(\widetilde{p_1}) \ge \logRL(p_1)$.  Similarly, $\logRL(\widetilde{p_2}) \ge \logRL(p_2)$ and $\logRL(\widetilde{p_3}) \ge \logRL(p_3)$.  That is, for points above and/or to the right of $B_1$, $\logRLss$ can be increased by moving down and/or to the left as far as $B_1$'s boundary.  Therefore \mrle\ must lie on or inside $B_1$ and there can be no maxima of $\logRLss$ in $B_1^c$.  $B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.  However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$.  This paper presents a computational method guaranteed not to miss \mrle\ and which evaluates $\logRLss$ everywhere to within a specified tolerance inside a box $B$ satisfying desideratum \textbf{D1}.  First, though, we pause to note that the region outlined in bold in Figure~(\ref{fig:smallboundingbox}) is a subset of $B_1$ that must also contain \mrle\ for the reasons given above.  But it is not rectangular, hence less convenient than $B_1$, so we don't pursue it further.

@@

$B$ is determined by its intercepts $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ on the $\sigesq$ and $\sigssq$ axes respectively.  Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis and let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$. Points such as $q_1$ which satisfy $\sigesq \le \sigma_e^{2*}$ and $\sigssq \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the left of the $s_z+1$ line --- are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are negative for $j \in \{1, \dots, s_z, s_z+2\}$ but nonnegative for $j=s_z+1$.  Let $\logRL_j(\cdot)$ denote the $j$'th term of \eqref{eq:sum} evaluated at $(\cdot)$.  
\begin{equation}
	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
\end{equation}
which is larger than $M$ if
\begin{equation}
\label{eq:q1}
  \begin{split}
    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
  \end{split}
\end{equation}
Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.  Thus $\widetilde{\sigma_s^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q1} is less than the r.h.s.~of \eqref{eq:q1}, and  \textbf{D1} is satisfied.

@@

Let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$.  For any point $q_4$ having $\sigssq \le \sigma_s^{2*}$ and $\sigesq \ge \widetilde{q_4}$, the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_z+1\}$  but nonnegative for $j=s_z+2$.  Therefore,
\begin{equation*}
	\logRL(\sshat) - \logRL(q_4) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
\end{equation*}
which is larger than $M$ if
\begin{equation}
\label{eq:q4}
  \begin{split}
    \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
      &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
      &= L - M + \frac{1}{2} \left[ c_{s_z+2}\log\sigma_s^{2*} + \frac{d_{s_z+2}}{\sigma_s^{2*}}\right].
  \end{split}
\end{equation}
Examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, and for $j \ne s_z+2$,\\ $\lim_{\sigesq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.  Thus $\widetilde{\sigma_e^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q4} is less than the r.h.s.~of \eqref{eq:q4}, and  \textbf{D1} is satisfied.

Combining the requirements for points like $q_1$, $q_2$, $q_3$, and $q_4$, desideratum \textbf{D1} is satisfied for all points in $B^c$ by choosing $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ sufficiently large. 
%enough to satisfy $\logRL(q_1^*) < L-M$ and choosing $\sigma_s^*$ large enough to satisfy $\logRL(q_2^*) < L-M$ and $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.  In addition, because $\logRL$ has no maxima outside of $B_1$, it also has no maxima outside of $B$.  As already mentioned, if there is an $h$ line, then $q_1$ would have to be treated like $q_2$ and $q_3$.

@@

\subsection{Lines Determine Bounds within Boxes}
\begin{figure}
	\centering
	\includegraphics[width=.5\linewidth]{figs/boxb.pdf}
	\caption{For a box $b$,  the minimum and
	maximum within $b$ of $\logRL_j$
	occur at either the lower left corner, the upper right  corner, or
	on the line.
}	\label{fig:boxb}
\end{figure}
Next we consider the relationship between the $\logRL_j$'s and boxes as depicted in Figure~\ref{fig:boxb}, which shows a box $b$ and three lines.  For lines such as $j=2$ that lie below or to the left of $b$, the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive, so the maximum and minimum of $\logRL_j$ within $b$ are attained at the lower left and upper right corners respectively.  The situation is reversed for lines like $j=3$ that lie above or to the right of $b$: the partial derivatives are nonnegative so the maximum and minimum are attained at the upper right and lower left corners respectively.  For lines like $j=1$ that pass through $b$, the minimum is attained at either the upper right or lower left corner while the maximum is attained on the line.

@@

\section{An Algorithm}
With \textbf{D1} and \textbf{D2} the following  algorithm will evaluate $\logRL$ to arbitrary accuracy everywhere within a box $B$.
\begin{enumerate}
\item Specify constants \maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$.
	\begin{enumerate}[label=(\alph*)]
	\item \maxit\ is the maximum number (could be $\infty$) of iterations of
		the algorithm's loop.
	\item We will not carefully analyze regions of the plane where
		$\logRLss < \logRL(\sshat) - M$.  ($M$ could be $\infty$.)
	\item We will evaluate $\logRL$ to within an accuracy of $\epsilon$ (could be 0) inside $B$
	         unless evaluation is stopped by one of the other criteria.
	\item We will not distinguish values of $\sigesq$ separated by less
		than $\delta_e$ (could be 0) nor values of $\sigssq$ separated by
		less than $\delta_s$ (could be 0).  Separation may be specified
		in either absolute or relative terms. (I.e.\ we look at either the difference
		in $\sigesq$ or $\log\sigesq$ (or $\sigssq$ or $\log\sigssq$) from one
		side of the box to the other.)
	\end{enumerate}
	\maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$ are constants that
	control the running of the algorithm.  Their meaning may become more
	clear as we explain the algorithm.
\item With $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ --- see \eqref{eq:lmm}
	and \eqref{eq:rll} for definitions and the Appendix for details --- compute $\{a_j, b_j, c_j, d_j\}$ for $j=1,
	\dots, s_z+2$.
\item Create two lists of boxes: active and inactive.  Both lists are initially empty.
\item Specify a box $B$ as the sole entry of the active list.  $B$ could be,
	but doesn't have to be, a box satisfying desideratum \textbf{D1}.
\item Either set $L = -\infty$ or evaluate $\logRL(p_i)$ at a few points
	$p_1, \dots p_k$ and set $L = \max_i\logRL(p_i)$.  $L$ is our current
	lower bound on $\logRL(\sshat)$.
\item While the active list is not empty or we have not reached \maxit\
	iterations
	\begin{enumerate}
	\item For each active box $b$:
		\begin{enumerate}
		\item if $U^b < L-M$ move $b$ to the inactive list.
		\item if $U^b - L^b < \epsilon$ move $b$ to the inactive list
		\item if the horizontal extent of $b$ is less than $\delta_e$,
			move $b$ to the inactive list
		\item if the vertical extent of $b$ is less than $\delta_s$,
			move $b$ to the inactive list
		\item otherwise, $b$ is still active.  Divide $b$ into four subboxes.
			Remove $b$ and add the subboxes to the active list.
		\end{enumerate}
	\item Set $L = \max ( L, \max_{b\in \text{active list}}L^b )$.  $L$ is our new
		lower bound on $\logRL(\sshat)$.
	\end{enumerate}
\item Return the active and inactive lists.
\end{enumerate}

@@

\subsubsection{A log RL Analysis}
For a $\logRLssorig$ analysis there are 43 lines, as shown in Figure~\ref{fig:hmolines}.
Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ axes with the settings
\begin{equation*}
	\maxit=10; \hspace{1cm} \epsilon=5; \hspace{1cm}
	\delta_e=\log(10); \hspace{1cm} \delta_s=\log(10); \hspace{1cm} M=5
\end{equation*}
results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 10: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$.  We see that boxes are steadily transferred from the active to the inactive list and that $L$ increases monotonically.  After 10 iterations there is a total of 640 boxes, which are displayed in Figure~\ref{fig:hmo1}.

@@

\begin{figure}
	\centering
	\includegraphics[width=.45\linewidth]{figs/hmo_HH11_lines.pdf}
	\caption{The 43 lines, $j=1, 2, \dots, 43$, for the $\logRLorig$ analysis of the HMO data.
	              The line for $j = s_z+1 = 43$ is dashed.}
	\label{fig:hmolines}
\end{figure}
